### 2013-03-01 article.tex

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps to required to
process raw sequence reads into a database-deposited genome sequence. Here we
describe ``Genomer,'' a tool that simplifies the manual tasks of finishing and
uploading a genome sequence to a database. Genomer can format a genome scaffold
into the common files required to submit to GenBank. Genomer simplifies editing
a genome scaffold by allowing an human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

@@

Manually editing a genome sequence or annotation is a non-trivial problem
requiring significant effort on a researcher's part. The large size of FASTA
sequence files or general feature format (GFF) annotation files makes changing
these files in a text editor difficult. Tasks such as organising project files,
adding additional sequences to an assembly, updating incorrect annotations or
preparing files for database deposition all divert effort from the main goal of
analysing a genome. As a result genome projects have a steep learning curve for
researchers embarking on genome sequencing for the first time, while still
requiring effort for experienced researchers.

We anticipate these problems have resulted in individual groups independently
producing in-house programming scripts to automate repetitive tasks. An example
of such tasks include finding protein names that do not match a convention of
three lower case characters followed by one upper case character (abcD). This
is manually laborious but simple to automate computationally. Individual
research groups producing their own scripts for such tasks however leads to a
repetition of effort. In contrast open-source projects can lead to a pooling of
community efforts resulting in a higher standard of software and preventing
\cite{ince2012}.

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer'', to ease and simplify many of the
common steps in a genome project. Several graphical user interface tools are
available to work with genome projects \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
line interface for genome finishing allowing scriptable automation and
reproducible results. We have open-sourced this tool under and MIT license and
here we describe its application for simplifying steps in a microbial genome
project.

@@

Genomer uses these two libraries to create a plugin system where third-party
software can be written and included. Plugin creators need only to prefix the
name of their plugin with \verb+genomer-plugin-+ and upload this to
rubygems.org or make the plugin available as a public git repository. A plugin
is specified within the project in a file named `Gemfile.' The \verb+bundle update+
command will then automatically download and install the required plugins.

Genomer is implemented as a command line tool and tested on both Mac OS X and
Linux systems. The latest version of genomer can installed from the Rubygems
package management system with the command \verb+gem install genomer+. Genomer
is then invoked on the command line with the command \verb+genomer+. Genomer
provides extensive documentation via man pages and on the genomer website at
http://next.gs.

@@

Genomer was developed during our \emph{P. fluorescens} genome projects
\cite{barton2012} to automate and simplify the manual steps required when
finishing a microbial genome. Genomer includes functions to simplify the
following: updating and modifying a genome sequence, editing genome
annotations, and preparing the required files to submit a draft or complete
genome to a sequence database such as GenBank. Genomer is also designed to
allow researchers to add their own functionality to a genome project through
third-party plugins.

This software was written for the command-line to use in shells such as bash or
zsh. Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. Additionally we stored each version of the GNU
Make build file with git \cite{git-scm} to track changes to the project and to
allow errors to be reverted from earlier versions. Example build files from our
genome projects can be found on GitHub for a simple plasmid
\cite{plasmid-github} and more complex \cite{genome-github} sequence.

@@

The Genomer plugin system was built using the existing Ruby package management
system: RubyGems and Bundler. This allows plugins to be automatically installed
without requiring manual downloading and compilation by the user. We anticipate
this will eliminate a common problem where software may require technical
expertise to first compile and install before being used. A plugin system also
allows each plugin to be developed on different release schedules without
requiring Genomer itself to be updated, thereby providing a stable interface.
Each Genomer plugin can be locked to specific versions to prevent changes in
functionality between different plugin versions. This addresses situations
whereby backwards incompatible software changes break an existing genomics when
a recent version is available.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
genomer may also be useful to others in the field. A practice described as
``eating your own dog food'' \cite{harrison2006}. We will illustrate the
possible usefulness of genomer with two use cases: updating the annotations of
a genome scaffold during the finishing process, and preparing genome files for
submission to GenBank.

@@

Generating a usable genome sequence requires sequencing, assembly of reads, and
annotation. A further step is also required: preparation of genomic files for
deposition in a database. An example is the GenBank database which requires the
genome annotations to be prepared in a custom table format. Additional files
are also required when the genome assembly is incomplete: a FASTA file
containing the individual contig sequences and `a golden path' (AGP)
\cite{agp-spec} file which describes the placement and direction of constituent
contigs in the draft genome sequence. The requirements to generate these
specific files can be time-consuming to complete and difficult to understand
for researchers new to field. We therefore Genomer during our own genome
projects to simplify and automate this process.

@@

Genomer does not produce the .asn which must be submitted to GenBank, instead
the generated files are used as the input for the tbl2asn or sequin tools. We
emphasise that the purpose of Genomer is not to reinvent the functionality
already provided by current NCBI resources, but instead to simplify the process
of producing the required inputs. Our example \emph{P. fluorescens} GNU Make
file [genome][] illustrates how Genomer may be combined with tbl2asn to produce
the required .asn file for submission.

>>>>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps to required to
process raw sequence reads into a database-deposited genome sequence. Here we
describe ``Genomer,'' a tool that simplifies the manual tasks of finishing and
uploading a genome sequence to a database. Genomer can format a genome scaffold
into the common files required to submit to GenBank. Genomer simplifies editing
a genome scaffold by allowing a human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

@@

Manually editing a genome sequence or annotation is a non-trivial problem
requiring significant effort on a researcher's part. The large size of FASTA
sequence files or general feature format (GFF) annotation files makes changing
these in a text editor difficult. Tasks such as organising project files,
adding additional sequences to an assembly, updating incorrect annotations or
preparing files for database deposition all divert effort from the main goal of
analysing a genome. As a result genome projects have a steep learning curve for
researchers embarking on genome sequencing for the first time, while still
requiring effort for experienced researchers.

We anticipate these problems have resulted in individual groups independently
producing in-house programming scripts to automate repetitive tasks. An example
of such tasks include finding protein names that do not match a convention of
three lower case characters followed by one upper case character (abcD). Such
tasks are manually laborious but simple to automate computationally. Individual
research groups producing their own scripts for such tasks however leads to a
repetition of effort. In contrast open-source projects can lead to a pooling of
community effort resulting in a higher standard of software and preventing
\cite{ince2012}.

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer'', to ease and simplify many of the
common steps in a genome project. Several graphical user interface tools are
available to work with genome projects \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
line interface for genome finishing allowing scriptable automation and
reproducible results. We have open-sourced this tool under the MIT license and
here we describe its application for simplifying steps in a microbial genome
project.

@@

Genomer uses these two libraries to create a plugin system where third-party
software can be written and included. Plugin creators need only to prefix the
name of their plugin with \verb+genomer-plugin-+ and upload this to
rubygems.org or make the plugin available as a public git repository. A plugin
is specified within a Genomer project in a file named `Gemfile.' The
\verb+bundle update+ command will then automatically download and install the
required plugins.

Genomer is implemented as a command line tool and tested on both Mac OS X and
Linux systems. The latest version of genomer can installed from the Rubygems
package management system with the command \verb+gem install genomer+. Genomer
is based on our previously described ``scaffolder'' software and is invoked on
the command line with the command \verb+genomer+. Genomer provides extensive
documentation via man pages and on the genomer website at http://next.gs.

@@

Genomer was developed during our \emph{P. fluorescens} genome projects to
automate and simplify the manual steps required when finishing a microbial
genome. Genomer includes functions to simplify the following: updating and
modifying a genome sequence, editing genome annotations, and preparing the
required files to submit a draft or complete genome to a sequence database such
as GenBank. Genomer is also designed to allow researchers to add their own
functionality to a genome project through third-party plugins.

This software was written for the command-line to use in shells such as bash or
zsh. Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. Additionally we stored each version of the GNU
Make build file with git \cite{git-scm} to track changes to the project and to
allow errors to be reverted from earlier versions. Example build files from our
genome projects can be found on GitHub for a simple plasmid
\cite{plasmid-github} and more complex genome \cite{genome-github} sequence.

@@

The Genomer plugin system was built using the existing Ruby package management
system: RubyGems and Bundler. This allows plugins to be automatically installed
without requiring manual downloading and compilation by the user. We anticipate
this will eliminate a common problem where software may require technical
expertise to first compile and install before being used. A plugin system also
allows each plugin to be developed on different release schedules without
requiring Genomer itself to be updated, thereby providing a stable interface.
Each Genomer plugin can be locked to specific versions to prevent changes in
functionality between different plugin versions. This addresses situations
whereby backwards incompatible software changes can break an existing project's
workflow.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
genomer may also be useful to others in the field. We will illustrate the
possible usefulness of genomer with two use cases: updating the annotations of
a genome scaffold during the finishing process, and preparing genome files for
submission to GenBank.

@@

Generating a usable genome sequence requires sequencing, assembly of reads, and
annotation. A further step is also required: preparation of genomic files for
deposition in a database. An example is the GenBank database which requires the
genome annotations to be prepared in a custom table format. Additional files
are also required when the genome assembly is incomplete: a FASTA file
containing the individual contig sequences and `a golden path' (AGP)
\cite{agp-spec} file which describes the placement and direction of constituent
contigs in the draft genome sequence. The requirements to generate these
specific files can be time-consuming to complete and difficult to understand
for researchers new to field. We used Genomer extensively during our own genome
projects to simplify and automate this process.

@@

Genomer does not produce the .asn file which must be submitted to GenBank,
instead the generated files are used as the input for the tbl2asn or sequin
tools. We emphasise that the purpose of Genomer is not to reinvent the
functionality already provided by current NCBI resources, but instead to
simplify the process of producing the required inputs. Our example \emph{P.
fluorescens} GNU Make file \cite{genome-github} illustrates how Genomer may be
combined with tbl2asn to produce the required .asn file for submission.
